---
title: "Final Project"
author: "Sanjay Renduchintala (vrenduch)"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: cerulean
    highlight: tango
---

### Problem Summary:

1. Develop an **attrition model**, to predict whether a customer will cancel their subscription in the near future. Characterize your model performance.
2. Develop a model for estimating the **ltv** of a customer. Characterize your model performance.
3. Develop a **customer segmentation** scheme. Include in this scheme the identification of sleeping customers, those that are no longer active but have not canceled their account.

#### About the dataset:
- id: A unique user identifier
- status: Subscription status:‘0’- new, ‘1’- open, ‘2’- cancelation event
- gender: User gender ‘M’- male ‘F’- female
- date: Date of in which user ‘id’ logged into the site
- pages: Number of pages visted by user ‘id’ on date ‘date’
- onsite: Number of minutes spent on site by user ‘id’ on date ‘date’
- entered: Flag indicating whether or not user entered the send order path on date ‘date’
- completed: Flag indicating whether the user completed the order (sent an eCard)
- holiday: Flag indicating whether at least one completed order included a holiday themed card

```{r}
# Import libraries
library(plyr)
library(ggplot2)
library(dplyr)
```

### 1. Exploratory Data Analysis

```{r}
ltv.data <- read.csv('ltv.csv')
head(ltv.data)
```

```{r}
ltv.data <- ltv.data %>% mutate_at(vars(id, status, gender, entered, completed, holiday), funs(as.factor(.)))
ltv.data$date <- as.Date(ltv.data$date, '%d/%m/%y')
summary(ltv.data)
```

#### 1.1. Missing values

```{r}
sapply(ltv.data, function(x) sum(is.na(x)))
```
**Observation:** There are no missing values in the data

#### 1.2. Extreme values

```{r}
hist(ltv.data$pages)
hist(ltv.data$onsite)
```
**Observation**: 'Onsite' has some extreme values

```{r}
# Quantiles
quantile(ltv.data$onsite, c(0,.1,.25,.5,.75,.9,.95,.99,1))
```

**Action:** Drop observations where user has spent more than 60 minutes on the website

#### 1.3. Summary of the data

```{r}
ltv.data <- ltv.data[ltv.data$onsite<=60,]
summary(ltv.data)
```


```{r}
# Number of users
print(paste0("Number of users: " , n_distinct(ltv.data$id)))

# Number of users by gender
print("Users by gender:")
print.data.frame(ltv.data %>% group_by(gender) %>% summarise(count = n_distinct(id)))

```

```{r}
# Attrition
att.summ <- ltv.data %>% mutate(month = format(date, "%m"), year = format(date, "%Y")) %>% group_by(year, status) %>% summarise(count = n_distinct(id))

att.summ$status <- mapvalues(att.summ$status, 
          from=c(0,1,2), 
          to=c("new","open","cancelled"))

ggplot(att.summ, aes(x = year, y = count,fill=status)) +
    geom_bar(stat='identity')
```

```{r}
# Life time value - top 10 customers with max. number of completed transactions
user_com <- ltv.data %>% filter(completed=="1") %>% group_by(id) %>% tally(sort = T)
head(user_com, 10)
```

### 2. Modeling


### 3. Key Findings



